Negative Perceptions of Outsourcing to Artificial Intelligence

### 2025-05-15 manuscript.qmd
Submit to JEPG

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
      - editing
  - name: Jim A.C. Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: This pre-print is currently not yet peer-reviewed, and may differ from the final version.
    data-sharing: Word count (excl. Methods and Results) = 4619 words.
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but 
  also socio-relational ones. Large language models like ChatGPT can generate 
  wedding vows, speeches, and personal messages, raising questions about how 
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (N = 3,649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks. We show
  that outsourcing makes us think more negatively about not only the person and
  their motivations, but also the outsourced work itself. Moreover, we provide
  insight into why this occurs: the reduced effort from outsourcing
  socio-relational tasks to AI signals that the output is less authentically 
  one's own and that the person cares less about the task. Our research 
  highlights the way that AI use shapes our perceptions of people, raising key
  philosophical questions about efficiency, authenticity, and social ties in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, person perception, outsourcing, effort, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf:
    numbered-lines: true
    include-in-header:
      text: \nolinenumbers
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

@@

The widespread release of generative AI language models has transformed daily 
life, offering the potential to perform a variety of tasks more efficiently and,
in some cases, with greater effectiveness than by doing them oneself. But as AI
becomes more widely available, people are not only using it to assist them with
things like preparing dinner recipes, writing data analysis code, and planning
daily schedules. Increasingly, AI might be used beyond routine or technical
domains to instead assist in tasks that are more socio-relational in nature,
like writing wedding vows, apology notes, and love letters. Anecdotal evidence
suggests that not only is AI-outsourcing of this kind already happening, but
that it potentially has serious effects on how we judge others. In a recent
Reddit post, a disgruntled newlywed tells the story of her husband using ChatGPT
to write his wedding vows, expressing her discomfort with outsourcing something
to AI that, to her, is deeply meaningful and a reflection of their love for one 
another [@miramar0]. Outsourcing tasks -- especially socio-relational ones -- to
AI tools may be efficient, but could have negative consequences for person 
perception.

>>>>

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
      - editing
  - name: Jim A.C. Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: This pre-print is currently not yet peer-reviewed, and may differ from the final version.
    data-sharing: Word count = 9119 words.
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but 
  also socio-relational ones. Large language models like ChatGPT can generate 
  wedding vows, speeches, and personal messages, raising questions about how 
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (N = 3,649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks. We show
  that outsourcing makes us think more negatively about not only the person and
  their motivations, but also the outsourced work itself. Moreover, we provide
  insight into why this occurs: the reduced effort from outsourcing
  socio-relational tasks to AI signals that the output is less authentically 
  one's own and that the person cares less about the task. Our research 
  highlights the way that AI use shapes our perceptions of people, raising key
  philosophical questions about efficiency, authenticity, and social ties in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, person perception, outsourcing, effort, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf:
    numbered-lines: true
    include-in-header:
      text: \nolinenumbers
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
impact-statement: >
  People form negative impressions of individuals who use artificial 
  intelligence tools to complete tasks for them, especially interpersonal tasks
  like writing love letters or apology notes. This occurs even when they use AI 
  as a tool instead of a replacement, are honest about it, and use AI because 
  they care about doing the task well. We show that doing something oneself, 
  even if AI could do it quicker and easier, signals that one is authentic and 
  cares about the task. Our results highlight that people do not only care about 
  whether something is done, but how it is done too.
language:
  title-impact-statement: "Significance Statement"
---

@@

The widespread release of generative AI language models has transformed daily 
life, offering the potential to perform a variety of tasks more efficiently and,
in some cases, with greater effectiveness than by doing them oneself. But as AI
becomes more widely available, people are not only using it to assist them with
things like preparing dinner recipes, writing data analysis code, and planning
daily schedules. Increasingly, AI might be used beyond routine or technical
domains to instead assist in tasks that are more socio-relational in nature,
like writing wedding vows, apology notes, and love letters. Anecdotal evidence
suggests that not only is AI-outsourcing of this kind already happening, but
that it potentially has serious effects on how we judge others. In a recent
Reddit post, a disgruntled newlywed tells the story of her husband using ChatGPT
to write his wedding vows, expressing her discomfort with outsourcing something
to AI that, to her, is deeply meaningful and a reflection of their love for one 
another [@miramar0]. Other reports describe people's negative reactions when
they learn that their romantic partner used ChatGPT to write them an apology 
note [@Tait2024] or even to break up with them [@Anderson2025]. Outsourcing 
tasks -- especially socio-relational ones -- to AI tools may be efficient, but 
could have negative consequences for person perception.
